---
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
The first code chunk contains variables that needs to be defined before 
running the code. 
```{r variables, include=FALSE}
library(tidyverse)
library(data.table)
library(scales)

### define variables ----------------------------------------------------------
# register(bpstart(SnowParam(2, progressbar = TRUE)))

# get working directory
wd <- dirname(getwd())

# set path to desired dataset
path <- paste0(wd, "/data/pathogen/")


# network file
net_df <- read.csv(paste0(path, "results/net_df.csv"))

# results file
results_top <- read.csv(paste0(path, "results/GPNAE_results_top.csv"))

# ids file containing ALL GPNAE Annotations
ids_fl <- paste0(
  path,  "ids.csv"
)

ids <- read.csv(ids_fl)[, 1:7]

colnames(ids) <- c(
  "Row", "m.z", "RT",
  "Annotation.ID", "formula",
  "adduct", "evidence"
)

# get all manual GPNAE species that will be used as seed node set (SNS)
gpnae <- ids[grepl("GPNAE", ids$Annotation.ID),]$Row



sirius_fl <- paste0(path, "SiriusComplete.csv")
```
## General Parameters and Statistics

Here is an overview about GPNAE network annotations by each layer: 

```{r eval, message=FALSE, warning=FALSE}
# no and type of seeds
length(gpnae)
gpnae
```
Next, we get an overview about the number of annotations that have been 
made with the algorithm.

```{r, include=FALSE}
## No of annotations
df_a <- table(results_top$network_layer) %>%
  as.data.frame()

df_a <- rbind(
  df_a,
  data.frame(Var1 = "total", Freq = sum(df_a$Freq))
)

colnames(df_a) <- c("layer", "n")

df_a$perc <- paste(
  round(100 * df_a$n / df_a[df_a$layer == "total", "n"],
    digits = 1
  ),
  "%"
)

df_a$layer <- factor(df_a$layer,
  levels = c(
    "total",
    "3_layers",
    "2_layers_gnps_mass",
    "2_layers_gnps_homol",
    "2_layers_mass_homol",
    "1_layer_gnps",
    "1_layer_mass"
  )
)
```

```{r}
# png(paste0(paste0(seeds_fold, "GPNAE_NoOfAnnotations.png")),
#     width = 800, height = 400, res = 100)

ggplot(data = df_a, aes(x = layer, y = n)) +
  geom_bar(stat = "identity", width = 0.7, fill = "#69005f") +
  geom_text(aes(label = perc), vjust = -0.3, size = 3) +
  labs(title = "No of Annotations") +
  theme_minimal() +
  scale_y_continuous(breaks = pretty_breaks(n = 12)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# dev.off()


# overview of number of network annotations:
table(results_top$network_layer)
```

## Validation using SIRIUS
The SIRIUS file is loaded  all its molecular formula are used and 
matched against all molecular formula that have been determined by the
network propagation algorithm.
If there is any match, the entry will be "yes", if there is none it will
be "no". If there is no molecular formula available, either in 
SIRIUS nor in the annotated species, the entry will be "NA". 
```{r, include=FALSE}
sirius <- read.csv(sirius_fl)

sirius <- select(sirius, Row, SIRIUS_molecularFormula)

sirius[sirius == ""] <- NA

results_top <- merge(results_top, sirius, by = "Row")


results_top$siriusMatch <- ifelse(sapply(
  lapply(results_top$network_formula, grepl,
    x = strsplit(results_top$SIRIUS_molecularFormula, split = "|", fixed = T)
  ),
  any
), "yes", "no")

results_top$siriusMatch <- ifelse(is.na(results_top$SIRIUS_molecularFormula), 
                                  NA, results_top$siriusMatch)


df_s <- table(results_top[, c("siriusMatch", "network_layer")],
  useNA = "always"
) %>% as.data.frame()

df_s <- rbind(df_s, data.frame(
  siriusMatch = "no", network_layer = "total",
  Freq = sum(df_s[df_s$siriusMatch == "no", ]$Freq, na.rm = T)
))


df_s <- rbind(df_s, data.frame(
  siriusMatch = "yes", network_layer = "total",
  Freq = sum(df_s[df_s$siriusMatch == "yes", ]$Freq, na.rm = T)
))


df_s <- rbind(df_s, data.frame(
  siriusMatch = NA, network_layer = "total",
  Freq = sum(df_s[is.na(df_s$siriusMatch), ]$Freq, na.rm = T)
))


df_s <- group_by(df_s, network_layer) %>% 
  mutate(perc = paste0((round(Freq / sum(Freq) * 100, 1)), " %"))

df_s <- df_s[!is.na(as.vector(df_s$network_layer)), ]

df_s$network_layer <- factor(df_s$network_layer,
  levels = c(
    "total", "3_layers", "2_layers_gnps_mass", "2_layers_gnps_homol",
    "2_layers_mass_homol", "1_layer_gnps", "1_layer_mass"
  )
)


# # save the results file with the SIRIUS results 
# write_csv(results_top,
#   file = paste0(seeds_fold, "sirius_results_top.csv")
# )

```
The first plot shows all matches including NA's, whereas the second plot 
removed the NA's, in order to better evaluate the matching annotations. 
The second plot provides the information of the percentage of features that 
have a matching and a miss-matching molecular formula (number of NA are not
considered.)

```{r sirius eval}
ggplot(data = df_s, aes(x = network_layer, y = Freq, fill = siriusMatch)) +
  geom_bar(stat = "identity", width = 0.7, position = "stack") +
  geom_text(aes(label = perc), position = position_stack(vjust = 0.5), size = 3.2) +
  labs(title = "No of Annotations") +
  theme_minimal() +
  scale_fill_manual(values = c("#69005f", "#ff506e", "black")) +
  scale_y_continuous(breaks = pretty_breaks(n = 12)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))


ggplot(data = df_s[!is.na(df_s$siriusMatch), ] %>% group_by(network_layer) %>% 
  mutate(perc = paste0((round(Freq / sum(Freq) * 100, 1)), " %")), 
  aes(x = network_layer, y = Freq, fill = siriusMatch)) +
  geom_bar(stat = "identity", width = 0.7, position = "stack") +
  geom_text(aes(label = perc), position = position_stack(vjust = 0.5), size = 3.2) +
  labs(title = "No of Annotations") +
  theme_minimal() +
  scale_fill_manual(values = c("#69005f", "#ff506e", "black")) +
  scale_y_continuous(breaks = pretty_breaks(n = 12)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

## Mass difference evaluation of annotated features

In this section we compare the observed mass-differences of 
1) the first neighbors of the gpnae seed nodes
2) between the seeds and the features that have been annotated
3) between the seeds and the features that not have been annotated
4) between the seeds and the features that have been annotated, but don't have
an spectral similarity edge

In order to get an overview of the appearance of mass-differences that are
more frequently used for the annotation of GPNAE's. 

```{r massdiff calc, include=FALSE}

## mass difference of first neighbors
tab_firstNeighbors <- table(net_df[net_df$from %in% gpnae |
  net_df$to %in% gpnae, ]$group_mass) %>%
  as.data.frame() %>%
  arrange(-Freq)



## get mass difference of annotated nodes
Annotated <- results_top[!is.na(results_top$network_annotation), ]$Row

## keep only edges with mass-differences
networkOfInterest <-
  net_df[((net_df$from %in% gpnae & net_df$to %in% Annotated |
    net_df$from %in% Annotated & net_df$to %in% gpnae) &
    !is.na(net_df$group_mass)), ]


tab_Annotated <- table(networkOfInterest$group_mass) %>%
  as.data.frame() %>%
  arrange(-Freq)


# no of mass difference annotations:
length(results_top[(!is.na(results_top$network_annotation) &
  results_top$network_layer == "1_layer_mass"), ]$Row)

## non annotated edges
NonAnnotated <- results_top[is.na(results_top$network_annotation), ]$Row

if (length(NonAnnotated) != 0) {
  networkOfInterest <-
  net_df[((net_df$from %in% gpnae & net_df$to %in% NonAnnotated |
    net_df$from %in% NonAnnotated & net_df$to %in% gpnae) &
    !is.na(net_df$group_mass)), ]

tab_NonAnnotated <- table(networkOfInterest$group_mass) %>%
  as.data.frame() %>%
  arrange(-Freq)
}



## annotated but w.o. spectral simil

AnnotatedNoSimilarity <- results_top[(!is.na(results_top$network_annotation) &
  results_top$network_layer != "1_layer_gnps"), ]$Row

networkOfInterest <-
  net_df[((net_df$from %in% gpnae & net_df$to %in% AnnotatedNoSimilarity |
    net_df$from %in% AnnotatedNoSimilarity & net_df$to %in% gpnae) &
    !is.na(net_df$group_mass) &
    is.na(net_df$gnps_cosine)
  ), ]

tab_AnnotatedNoSimil <- table(networkOfInterest$group_mass) %>%
  as.data.frame() %>%
  arrange(-Freq)



## annotated with correct SIRIUS
## get mass difference of annotated nodes
Annotated_sir_y <- results_top[!is.na(results_top$siriusMatch) &
                                 results_top$siriusMatch == "yes", ]$Row

## keep only edges with matching sirius
networkOfInterest <-
  net_df[((net_df$from %in% gpnae & net_df$to %in% Annotated_sir_y  |
    net_df$from %in% Annotated_sir_y  & net_df$to %in% gpnae) &
    !is.na(net_df$group_mass)
    ), ]


tab_Annotated_sirY <- table(networkOfInterest$group_mass) %>%
  as.data.frame() %>%
  arrange(-Freq)


## annotated and wrong SIRIUS
## get mass difference of annotated nodes
Annotated_sir_n <- results_top[!is.na(results_top$siriusMatch) &
                                 results_top$siriusMatch == "no", ]$Row

## keep only edges with matching sirius
networkOfInterest <-
  net_df[((net_df$from %in% gpnae & net_df$to %in% Annotated_sir_n  |
    net_df$from %in% Annotated_sir_n  & net_df$to %in% gpnae) &
    !is.na(net_df$group_mass)
    ), ]


tab_Annotated_sirN <- table(networkOfInterest$group_mass) %>%
  as.data.frame() %>%
  arrange(-Freq)



tab_master <- merge(tab_Annotated, tab_AnnotatedNoSimil, by = "Var1", all = TRUE)
colnames(tab_master) <- c("Var1", "Annotated", "AnnotatedNoSimil")

tab_master <- merge(tab_master, tab_firstNeighbors, by = "Var1", all = TRUE)
colnames(tab_master) <- c(colnames(tab_master)[-length(tab_master)], "firstNeighbors")

tab_master <- merge(tab_master, tab_Annotated_sirY, by = "Var1", all = TRUE)
colnames(tab_master) <- c(colnames(tab_master)[-length(tab_master)], "siriusMatch")

if (nrow(tab_Annotated_sirN) != 0) {
tab_master <- merge(tab_master, tab_Annotated_sirN, by = "Var1", all = TRUE)
colnames(tab_master) <- c(colnames(tab_master)[-length(tab_master)], "siriusMissMatch")
}
if (length(NonAnnotated) != 0) {
 tab_master <- merge(tab_master, tab_NonAnnotated, by = "Var1", all = TRUE)
 colnames(tab_master) <- c(colnames(tab_master)[-length(tab_master)], "NonAnnotated")
}

# write_csv(tab_master,
#   file = paste0(seeds_fold, "MassDifferencesStatistics.csv")
# )

```
Summary of mass differences of:
```{r mass-diff}
# first neighbors of seeds
head(tab_firstNeighbors)
# Annotated nodes 
head(tab_Annotated)

# NOT Annotated nodes 
if (length(NonAnnotated) != 0) {
head(tab_NonAnnotated)
}
# Annotated nodes but without spectral similarity
head(tab_AnnotatedNoSimil)
# Annotated nodes that match to SIRIUS molecular formula
head(tab_Annotated_sirY)
# Annotated nodes that NOT match to SIRIUS  molecular formula
head(tab_Annotated_sirN)

```

## Glucose annotations
The annotation of Glucose variants of GPNAEs is mainly accomplished by 
propagation in die mass-difference layer because the MS2 spectra are 
too different. 
Therefore, this section evaluates the properties of the Glucose annotations of this dataset. 
First, we want to create a sub-network and keep glucose annotations only.
```{r Gluc}
annoGluc <- results_top[results_top$network_annotation %like% "C6H10O5", ]

#Number of Glucose annotations
nrow(annoGluc)

#Number of ms2 available at Glucose annotations
table(annoGluc$ms2_available)

#Layer that have been used for Glucose annotation
table(annoGluc$network_layer)

## how many with sirius correct 
table(annoGluc$siriusMatch)

```

```{r, include=FALSE}
#Keep only seed nodes and glucose annotations and all their edges 
glu_net <- net_df[net_df$from %in% annoGluc$Row | net_df$to %in% annoGluc$Row |
  net_df$from %in% gpnae | net_df$to %in% gpnae, ]

# Keep only edges between Glucose annotations and seed nodes
glu_edge <- glu_net[glu_net$from %in% gpnae & glu_net$to %in% annoGluc$Row |
  glu_net$to %in% gpnae & glu_net$from %in% annoGluc$Row, ]

# edges withing glucose subnetwork
inner_glu <-
  glu_net[glu_net$from %in% annoGluc$Row &
    glu_net$to %in% annoGluc$Row, ]
```
how many spec edges within glucose sub network and their mean 
spectral similarity

```{r}
# number of glucose annotations that have spectral similarity within each other
length(inner_glu[!is.na(inner_glu$gnps_cosine), ]$gnps_cosine)
# mean spectral similarity
mean(inner_glu[!is.na(inner_glu$gnps_cosine), ]$gnps_cosine)
```

### Evaluation of benefits of Homol Series

We test if we see any benefits/ better annotation by features that are 
somehow connected in the homologous series network, because they are more 
likely to be real biological features because of the RT trends. 
Therefore we create a sub network, where we remove all features that show 
no connection in the homologous series layer. 


```{r }
# remove nodes that have no homol series (from original file)
homol_net <- net_df[!is.na(net_df$HSIDs), ]

#get cluster_ID of the nodes that appear on homol series layer
homols <- unique(c(homol_net$from, homol_net$to))

# filter results file by those features appearing in homol series
results_homol <- results_top[results_top$Row %in% homols, ]

    
## how many left
nrow(results_homol)
nrow(results_top)

#Layer that have been used for annotation
table(results_homol$network_layer)
table(results_top$network_layer)

## how many with MS2
table(results_homol$ms2_available)
table(results_top$ms2_available)

## how many with sirius correct compared to all nodes
table(results_homol$siriusMatch)
table(results_top$siriusMatch)

```

#### Evaluation using Glucose Annotations
```{r}
## create subnetwork with Glucose nodes and nodes that appeat in homol series
annoGluc_homol <- results_homol[results_homol$network_annotation %like% "C6H10O5", ]

## how many left
nrow(annoGluc_homol)

#Layer that have been used for Glucose annotation
table(annoGluc_homol$network_layer)

## how many with MS2
table(annoGluc_homol$ms2_available)

## how many with sirius correct 
table(annoGluc_homol$siriusMatch)


```
## Validation using manually Annotated nodes

In order to verify how good annotations are with the network propagation algorithm, we next create a sub network were we only keep nodes where we have an manual annotation (i.e. a molecular formula for matching). 
With that we can first validate how good our annotations are and secondly, validate if filtering nodes for the homologous series increases our annotation success or if we just lose important annotations. 

```{r, include=FALSE}
# keep only ids where molecular formula is available
validMol <- ids[ids$formula != "", ]

validMol <- select(validMol, Row, Annotation.ID, molFormula = formula)
validMol[validMol == ""] <- NA


results_valid <- merge(results_top, validMol, by = "Row", all.x = TRUE)

results_valid <- results_valid %>% 
  select(Row, `m.z`, RT, "Annotation.ID.x", "formula", 
         "ms2_available", "network_annotation", "network_formula",
         "Annotation.ID.y","molFormula",
         "network_inputNode", "network_layer",
         "network_comment", "key_fragments"
         )


results_valid$validMolMatch <- ifelse(sapply(
  lapply(results_valid$network_formula, grepl,
    x = results_valid$molFormula
  ),
  any
), "yes", "no")

results_valid$validMolMatch <- ifelse(is.na(results_valid$molFormula), 
                                  NA,results_valid$validMolMatch)







noMatch <- results_valid[results_valid$validMolMatch == "no" & 
                           !is.na(results_valid$validMolMatch), ]

Match <- results_valid[results_valid$validMolMatch == "yes" & 
                           !is.na(results_valid$validMolMatch), ]

```

```{r}
#how many does match 
nrow(Match)

#how many does NOT match
nrow(noMatch)
#comment on network to see what e.g. mass differences can be excluded
noMatch$network_comment


```
### Test if homol series increase the success 




```{r}
## test how many homols have key fragments 
table(select(results_top, homol, key_fragments))

## test how many homols in 2_layer annotation gnps+mass
table(results_top[results_top$network_layer == "2_layers_gnps_mass", ]$homol)

## test how many homols in 1_layer annotation gnps
table(results_top[results_top$network_layer == "1_layer_gnps", ]$homol)



homol_valid <- merge(results_homol, validMol, by = "Row", all.x = TRUE)

homol_valid <- homol_valid %>% 
  select(Row, `m.z`, RT, "Annotation.ID.x", "formula", 
         "ms2_available", "network_annotation", "network_formula",
         "Annotation.ID.y","molFormula",
         "network_inputNode", "network_layer",
         "network_comment", "key_fragments"
         )


homol_valid$validMolMatch <- ifelse(sapply(
  lapply(homol_valid$network_formula, grepl,
    x = homol_valid$molFormula
  ),
  any
), "yes", "no")

homol_valid$validMolMatch <- ifelse(is.na(homol_valid$molFormula), 
                                  NA,homol_valid$validMolMatch)


noMatchHomol <- homol_valid[homol_valid$validMolMatch == "no" & 
                           !is.na(homol_valid$validMolMatch), ]

MatchHomol <- homol_valid[homol_valid$validMolMatch == "yes" & 
                           !is.na(homol_valid$validMolMatch), ]


#how many does match 
nrow(MatchHomol)

#how many does NOT match
nrow(noMatchHomol)
#comment on network to see what e.g. mass differences can be excluded
noMatchHomol$network_comment

## get differences between matches of homol series 
## filtered results file and normal file


diff <- anti_join( Match, MatchHomol)
# All those annotations are missed when using homol series to filter all nodes:
diff$Annotation.ID.y

# check which layers are affected

diff$network_layer




```





```{r, include=FALSE}

df_s <- table(results_homol[, c("siriusMatch", "network_layer")],
  useNA = "always"
) %>% as.data.frame()

df_s <- rbind(df_s, data.frame(
  siriusMatch = "no", network_layer = "total",
  Freq = sum(df_s[df_s$siriusMatch == "no", ]$Freq, na.rm = T)
))


df_s <- rbind(df_s, data.frame(
  siriusMatch = "yes", network_layer = "total",
  Freq = sum(df_s[df_s$siriusMatch == "yes", ]$Freq, na.rm = T)
))


df_s <- rbind(df_s, data.frame(
  siriusMatch = NA, network_layer = "total",
  Freq = sum(df_s[is.na(df_s$siriusMatch), ]$Freq, na.rm = T)
))


df_s <- group_by(df_s, network_layer) %>% 
  mutate(perc = paste0((round(Freq / sum(Freq) * 100, 1)), " %"))

df_s <- df_s[!is.na(as.vector(df_s$network_layer)), ]

df_s$network_layer <- factor(df_s$network_layer,
  levels = c(
    "total", "3_layers", "2_layers_gnps_mass", "2_layers_gnps_homol",
    "2_layers_mass_homol", "1_layer_gnps", "1_layer_mass"
  )
)
```
#### plot homols 




```{r}
## SIRIUS matches when molecules are part of homol series 

ggplot(data = df_s, aes(x = network_layer, y = Freq, fill = siriusMatch)) +
  geom_bar(stat = "identity", width = 0.7, position = "stack") +
  geom_text(aes(label = perc), position = position_stack(vjust = 0.5), size = 3.2) +
  labs(title = "No of Annotations") +
  theme_minimal() +
  scale_fill_manual(values = c("#69005f", "#ff506e", "black")) +
  scale_y_continuous(breaks = pretty_breaks(n = 12)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))


ggplot(data = df_s[!is.na(df_s$siriusMatch), ] %>% group_by(network_layer) %>% 
  mutate(perc = paste0((round(Freq / sum(Freq) * 100, 1)), " %")), 
  aes(x = network_layer, y = Freq, fill = siriusMatch)) +
  geom_bar(stat = "identity", width = 0.7, position = "stack") +
  geom_text(aes(label = perc), position = position_stack(vjust = 0.5), size = 3.2) +
  labs(title = "No of Annotations") +
  theme_minimal() +
  scale_fill_manual(values = c("#69005f", "#ff506e", "black")) +
  scale_y_continuous(breaks = pretty_breaks(n = 12)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))


```

#### plot results_top
1_layer_mass is now filtered: Only those nodes are kept that show any homol series 






```{r, include=FALSE}

# remove nodes that have no homol series from 1_layer_mass
df_s <- table(results_top[!(results_top$homol == "no" &
              results_top$network_layer == "1_layer_mass"), 
              c("siriusMatch", "network_layer")],
  useNA = "always"
) %>% as.data.frame()

df_s <- rbind(df_s, data.frame(
  siriusMatch = "no", network_layer = "total",
  Freq = sum(df_s[df_s$siriusMatch == "no", ]$Freq, na.rm = T)
))


df_s <- rbind(df_s, data.frame(
  siriusMatch = "yes", network_layer = "total",
  Freq = sum(df_s[df_s$siriusMatch == "yes", ]$Freq, na.rm = T)
))


df_s <- rbind(df_s, data.frame(
  siriusMatch = NA, network_layer = "total",
  Freq = sum(df_s[is.na(df_s$siriusMatch), ]$Freq, na.rm = T)
))


df_s <- group_by(df_s, network_layer) %>% 
  mutate(perc = paste0((round(Freq / sum(Freq) * 100, 1)), " %"))

df_s <- df_s[!is.na(as.vector(df_s$network_layer)), ]

df_s$network_layer <- factor(df_s$network_layer,
  levels = c(
    "total", "3_layers", "2_layers_gnps_mass", "2_layers_gnps_homol",
    "2_layers_mass_homol", "1_layer_gnps", "1_layer_mass"
  )
)
```




```{r}
## SIRIUS matches when molecules are part of homol series 

ggplot(data = df_s, aes(x = network_layer, y = Freq, fill = siriusMatch)) +
  geom_bar(stat = "identity", width = 0.7, position = "stack") +
  geom_text(aes(label = perc), position = position_stack(vjust = 0.5), size = 3.2) +
  labs(title = "No of Annotations") +
  theme_minimal() +
  scale_fill_manual(values = c("#69005f", "#ff506e", "black")) +
  scale_y_continuous(breaks = pretty_breaks(n = 12)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))


table(results_top[!(results_top$homol == "no" &
              results_top$network_layer == "1_layer_mass"), 
              "network_layer"]
)

table(results_top[, 
              "network_layer"]
)


```


## Annotated as LPEs 


```{r}
## SIRIUS matches when molecules are part of homol series 

LPEs <- results_top[grepl("LPE", results_top$Annotation.ID),]

# no of LPEs that were annotates as GPNAEs in the APEX
nrow(LPEs)

# layers that have been used
table(LPEs$network_layer
)

# how many part of homol series
table(LPEs$homol)



```





